---
title: "Performance Trends in the 2025 Connecticut Sun Season"
author: "Morde Bernstein, Marley McGehee, Kimberly McNeil"
date: "4/17/2026"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) #add error = FALSE and message = FALSE when complete!!!
library(tidyverse)
library(lubridate)
library(knitr)
```

## Introduction

Our dataset is the 2025 schedule and results for the Connecticut Sun, a WNBA team. The data looks at the dates when the games were played, who they played against, the scores, as well as whether they were on a win/lose streak. Additional factors include where they played the game, if it went into overtime, and any additional notes. The sample size is 44. We obtained the data from Basketball Reference's website on the [2025 Connecticut Sun Schedule and Results](http://basketball-reference.com/wnba/teams/CON/2025_games.html) page.

## Data Import

```{r}
basketball <- read_csv("connecticut_sun_2025.csv", 
                       col_types = cols(Date = col_date("%a %b %d %Y")))

# Initialize cleaning the column names (and giving names to the unnamed)
basketball <- janitor::clean_names(basketball) 

# Make the janitor's work clearer
basketball <- basketball %>%
  rename(
    game = g,
    home_or_away = x3,
    result = x5,
    overtime = x6,
    team_points = tm,
    opp_points = opp,
    total_wins = w,
    total_losses = l
  ) %>%
  mutate(
    home_or_away = ifelse(is.na(home_or_away) | home_or_away == "", "Home", "Away"),
    overtime = ifelse(is.na(overtime) | overtime == "", "False", "True")
)

```

We first imported the data as a .csv file, then used a function to make the date variable be recognized as date data instead of simply text. While plain text is easy for humans to read, computers can have difficulty, and this will allow it to understand that it is a date and be able to validate our data (e.g. dates like Feburary 30th will be flagged and denied) and perform unique calculations (e.g. timespans). When compared to most datasets with dates, this dataset uniquely contains an abbreviated day of the week in the date variable, but we decided not to include it in our current table because later functions will be able to pull the day of the week from the date alone, so it is not necessary. Following this, we cleaned the names of the columns and updated missing/confusing labels so they would be easier to both use and understand going forward. 

Let's take a brief look at the data:

```{r}
kable(basketball %>%
        head(n = 5), 
      caption = "The First Five Rows of our Cleaned Basketball Data")
```


## Exploratory Data Analysis

### Variable Exploration

#### Date Variable

```{r}
basketball %>%
  mutate(wday = wday(date, label = TRUE)) %>%
  ggplot(aes(x = wday)) +
  geom_bar(color = "black",
           fill = "darkorange") +
  labs (x = "Day of the Week",
        title = "Day of the Week Breakdown")
```

Here we can see the breakdown of days of the week in which games were played. Games were played on all days of the week with Sunday being by far the most popular day, and Thursday being the least popular. 

```{r}
basketball %>%
  mutate(month = month(date, label = TRUE)) %>%
  ggplot(aes(x = month)) +
  geom_bar(color = "black",
           fill = "darkorange") +
  labs(title = "Month Data")
```

Here we can see the breakdown of months in which games were played. Games were played from May to September, with August having the most games played in it. This makes sense as the WNBA season starts in May and ends in September.


```{r}
ggplot(basketball %>%
         mutate(day_type = ifelse(wday(date) %in% c(1,7), "Weekend", "Weekday")),
       aes(x = day_type)) +
  geom_bar(fill = "darkorange") +
  labs(title = "Number of Games on Weekends vs Weekdays",
       x = "Weekday or Weekend")
```

Here we can see that they played more games on weekdays than on weekends. This could be useful for later analysis as there is more data for games played on weekdays. 


#### Team Score

```{r}
basketball %>%
  summarize(
    Count = n(),
      Mean = mean(team_points, na.rm = TRUE),
      Median = median(team_points, na.rm = TRUE),
      sd = sd(team_points, na.rm = TRUE),
      Min = min(team_points, na.rm = TRUE),
      Q1 = quantile(team_points, 0.25, na.rm = TRUE),
      Q3 = quantile(team_points, 0.75, na.rm = TRUE),
      Max = max(team_points, na.rm = TRUE)
  ) %>% 
  kable(caption = "Summary Statistics for Team Points")
```

```{r}
basketball %>% 
  ggplot(aes(x = team_points)) +
    geom_histogram(bins = 15,
                   fill = "darkorange",
                   color = "black") +
    labs(title = "Distribution of Team Points (Histogram)",
         x = "Team Score",
         y = "Count")
```



```{r}
ggplot(basketball, aes(x=team_points))+
  geom_boxplot(fill = "darkorange", color = "black") +
  labs (title = "Team Score Boxplot", 
    x = "Team Score")

```

Above we can see visualizations of the team score per game. Across all 44 of the games, the lowest score was 52 and the highest score was 101. The average score was approximately 75 pts. 


#### Bivariate Relationship
```{r}
ggplot(basketball, aes(x = date, y = team_points)) +
  geom_point(alpha = 0.2)+ 
  geom_smooth(color = "darkorange") +
  labs(title = "Team Scores Over Time", 
       x = "Date", 
       y = "Team Score")
```

Here we can see the relationship between the date and team score. There appears to be some fluctuation due to month, with scores increasing in August, and then decreasing into September. 


### Time-Based Patterns

#### Day/Month 

The following summary and bar plot showcase the Average Team Score by the Day of the Week: 


```{r}
basketball %>%
  mutate(day = wday(date, label = TRUE)) %>%
  group_by(day) %>%
  summarize(avg_score = mean(team_points)) %>%
  kable(caption = "Average Team Score by Day of the Week")
```


```{r}
basketball %>%
  mutate(day = wday(date, label = TRUE)) %>%
  group_by(day) %>%
  summarize(avg_score = mean(team_points)) %>%
  ggplot(aes(x = day, y = avg_score)) +
  geom_col(fill = "darkorange", color = "black") +
  labs(title = "Average Team Score by Day of the Week",
       x = "Day of the Week",
       y = "Average Team Score")
```

As shown above, the average team score is lowest on Monday and highest on Saturday. Sundays, the weekday with the most games by plurality, have the second lowest average team score.


```{r}
basketball %>%
  mutate(month = month(date, label = TRUE)) %>%
  group_by(month) %>%
  summarize(avg_score = mean(team_points)) %>%
  kable(caption = "Average Team Score by Month")
```


```{r}
basketball %>%
  mutate(month = month(date, label = TRUE)) %>%
  group_by(month) %>%
  summarize(avg_score = mean(team_points)) %>%
  ggplot(aes(x = month, y = avg_score)) +
  geom_col(fill = "darkorange", color = "black") +
  labs(title = "Average Team Score by Month",
       x = "Month",
       y = "Average Team Score")
```

June's average team score was the lowest at 69.5, but immediately after, July had the highest average team score of 82.1. The Suns' average team score gradually decreased going into August and later September.


#### Weekend vs Weekday

```{r}
basketball %>%
  mutate(day_type = ifelse(wday(date) %in% c(1,7), "Weekend", "Weekday")) %>% # 1 = Sun and 7 = Sat
  group_by(day_type) %>%
  summarize(avg_score = mean(team_points)) %>%
  kable(caption = "Average Team Scores on Weekdays vs Weekends")
```

```{r}
ggplot(basketball %>%
         mutate(day_type = ifelse(wday(date) %in% c(1,7), "Weekend", "Weekday")),
       aes(x = day_type, y = team_points)) +
  geom_boxplot(fill = "darkorange") +
  labs(title = "Average Team Scores on Weekdays vs Weekends",
       x = "Weekday or Weekend", 
       y = "Average Team Score")
```

The Suns tended to perform better on weekdays, with a higher minimum and maximum average team score, and a higher median average team score than on weekends. Since we know that fewer games were played on weekends, it could be possible that the apparent higher inter-quartile range is because that smaller number of games means the quartiles have stretched slightly.  


#### Season/Quarter


```{r}
basketball %>%
  mutate(quarter = quarter(date)) %>%
  group_by(quarter) %>%
  summarize(avg_score = mean(team_points)) %>%
  kable(caption = "Average Team Scores by Quarter")
```

```{r}
basketball %>%
  mutate(quarter = quarter(date)) %>%
  group_by(quarter) %>%
  summarize(avg_score = mean(team_points)) %>%
  ggplot(aes(x = factor(quarter), y = avg_score)) +
  geom_col(fill = "darkorange", color = "black") +
  labs(title = "Average Team Score by Season Quarter",
       x = "Quarter",
       y = "Average Team Score")
```


The Connecticut Suns played games in quarters 2 and 3 with quarter comprising April through June and quarter 3 being from July to September. The team score was slightly higher in quarter 3 with an average of 78 points compared to 71 points in quarter 2. 


## Conclusion

