---
title: "Performance Trends in the 2025 Connecticut Sun Season"
author: "Morde Bernstein, Marley McGehee, Kimberly McNeil"
date: "4/17/2026"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) #add error = FALSE and message = FALSE when complete!!!
library(tidyverse)
library(lubridate)
library(knitr)
```

## Introduction

Our dataset is the 2025 schedule and results for the Connecticut Sun, a WNBA team. The data looks at the dates when the games were played, who they played against, the scores, as well as whether they were on a win/lose streak. Additional factors include where they played the game, if it went into overtime, and any additional notes. The sample size is 44. We obtained the data from Basketball Reference's website on the [2025 Connecticut Sun Schedule and Results](http://basketball-reference.com/wnba/teams/CON/2025_games.html) page.

## Data Import

```{r}
basketball <- read_csv("connecticut_sun_2025.csv", 
                       col_types = cols(Date = col_date("%a %b %d %Y")))

# Initialize cleaning the column names (and giving names to the unnamed)
basketball <- janitor::clean_names(basketball) 

# Make the janitor's work clearer
basketball <- basketball %>%
  rename(
    game = g,
    home_or_away = x3,
    result = x5,
    overtime = x6,
    team_points = tm,
    opp_points = opp,
    total_wins = w,
    total_losses = l
  ) %>%
  mutate(
    home_or_away = ifelse(is.na(home_or_away) | home_or_away == "", "Home", "Away"),
    overtime = ifelse(is.na(overtime) | overtime == "", "False", "True")
)

```

We first imported the data as a .csv file, then used a function to make the date variable be recognized as date data instead of simply text. While plain text is easy for humans to read, computers can have difficulty, and this will allow it to understand that it is a date and be able to validate our data (e.g. dates like Feburary 30th will be flagged and denied) and perform unique calculations (e.g. timespans). When compared to most datasets with dates, this dataset uniquely contains an abbreviated day of the week in the date variable, but we decided not to include it in our current table because later functions will be able to pull the day of the week from the date alone, so it is not necessary. Following this, we cleaned the names of the columns and updated missing/confusing labels so they would be easier to both use and understand going forward. 

Let's take a brief look at the data:

```{r}
kable(basketball %>%
        head(n = 5), 
      caption = "The First Five Rows of our Cleaned Basketball Data")
```


## Exploratory Data Analysis

### Variable Exploration

#### Date Variable

```{r}
basketball %>%
mutate(wday = wday(date, label = TRUE)) %>%
ggplot(aes(x = wday)) +
geom_bar(color = "black",
         fill = "darkorange") +
  labs (x = "Day of the Week")
```

```{r}
basketball %>%
mutate(month = month(date, label = TRUE)) %>%
ggplot(aes(x = month)) +
geom_bar(color = "black",
         fill = "darkorange") 
```


#### Team Score

```{r}
basketball %>%
  summarize(
    Count = n(),
      Mean = mean(team_points, na.rm = TRUE),
      Median = median(team_points, na.rm = TRUE),
      sd = sd(team_points, na.rm = TRUE),
      Min = min(team_points, na.rm = TRUE),
      Q1 = quantile(team_points, 0.25, na.rm = TRUE),
      Q3 = quantile(team_points, 0.75, na.rm = TRUE),
      Max = max(team_points, na.rm = TRUE)
  ) %>% 
  kable(caption = "Summary Statistics for Team Points")
```

```{r}
basketball %>% 
  ggplot(aes(x = team_points)) +
    geom_histogram(bins = 15,
                   fill = "darkorange",
                   color = "black") +
    labs(title = "Distribution of Team Points (Histogram)",
         x = "Team Score",
         y = "Count")
```
```{r}
ggplot(basketball, aes(x=team_points))+
  geom_boxplot() +
  labs (title = "Team Score Boxplot", 
    x = "Team Score")

```

#### Bivariate

```{r}
ggplot(basketball, aes(x = date, y = team_points)) +
  geom_point(alpha = 0.2)+ 
  geom_smooth(color = "darkorange") +
  labs(x = "Date", y = "Team Score")
```



