---
title: "Project 2"
author: "Morde Bernstein, Marley McGehee, Kimberly McNeil"
date: "4/17/2026"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(knitr)
```

## Introduction

Our dataset is the 2025 schedule and results for the Connecticut Sun, a WNBA team. The data looks at the dates when the games were played, who they played against, the scores, as well as whether they were on a win/lose streak. Additional factors included were where they played the game, if it went into overtime, and additional notes. The sample size was 44. The data was obtained from basketball reference at this link http://basketball-reference.com/wnba/teams/CON/2025_games.html. 

## Data Import

```{r}
basketball <- read_csv("
G,Date,,Opponent,,,Tm,Opp,W,L,Streak,Notes
1,Sun May 18 2025,,Washington Mystics,L,,85,90,0,1,L 1,
2,Tue May 20 2025,,Las Vegas Aces,L,,62,87,0,2,L 2,
3,Fri May 23 2025,@,Minnesota Lynx,L,,70,76,0,3,L 3,
4,Sun May 25 2025,@,Atlanta Dream,L,,55,79,0,4,L 4,
5,Tue May 27 2025,,Dallas Wings,L,,87,109,0,5,L 5,
6,Fri May 30 2025,@,Indiana Fever,W,,85,83,1,5,W 1,
7,Sun Jun 1 2025,@,New York Liberty,L,,52,100,1,6,L 1,Commissioner's Cup Game
8,Fri Jun 6 2025,,Atlanta Dream,W,,84,76,2,6,W 1,Commissioner's Cup Game
9,Sun Jun 8 2025,@,Washington Mystics,L,,67,104,2,7,L 1,Commissioner's Cup Game
10,Sun Jun 15 2025,,Chicago Sky,L,,66,78,2,8,L 2,Commissioner's Cup Game
11,Tue Jun 17 2025,@,Indiana Fever,L,,71,88,2,9,L 3,Commissioner's Cup Game
12,Wed Jun 18 2025,,Phoenix Mercury,L,,75,83,2,10,L 4,
13,Fri Jun 20 2025,,Dallas Wings,L,,83,86,2,11,L 5,
14,Sun Jun 22 2025,@,Golden State Valkyries,L,,63,87,2,12,L 6,
15,Wed Jun 25 2025,@,Las Vegas Aces,L,,59,85,2,13,L 7,
16,Fri Jun 27 2025,@,Seattle Storm,L,,81,97,2,14,L 8,
17,Sun Jun 29 2025,@,Minnesota Lynx,L,,63,102,2,15,L 9,
18,Sun Jul 6 2025,,Las Vegas Aces,L,,68,86,2,16,L 10,
19,Wed Jul 9 2025,,Seattle Storm,W,,93,83,3,16,W 1,
20,Fri Jul 11 2025,@,Seattle Storm,L,,65,79,3,17,L 1,
21,Sun Jul 13 2025,@,Los Angeles Sparks,L,,88,92,3,18,L 2,
22,Tue Jul 15 2025,,Indiana Fever,L,,77,85,3,19,L 3,
23,Thu Jul 24 2025,,Los Angeles Sparks,L,,86,101,3,20,L 4,
24,Sun Jul 27 2025,,Golden State Valkyries,W,,95,64,4,20,W 1,
25,Mon Jul 28 2025,,Seattle Storm,L,,85,101,4,21,L 1,
26,Fri Aug 1 2025,,New York Liberty,W,,78,62,5,21,W 1,
27,Sun Aug 3 2025,,New York Liberty,L,,78,87,5,22,L 1,
28,Tue Aug 5 2025,@,Phoenix Mercury,L,,66,82,5,23,L 2,
29,Thu Aug 7 2025,@,Los Angeles Sparks,L,,91,102,5,24,L 3,
30,Sun Aug 10 2025,@,Las Vegas Aces,L,,86,94,5,25,L 4,
31,Mon Aug 11 2025,@,Golden State Valkyries,L,,57,74,5,26,L 5,
32,Wed Aug 13 2025,,Chicago Sky,W,,71,62,6,26,W 1,
33,Sun Aug 17 2025,,Indiana Fever,L,OT,93,99,6,27,L 1,
34,Tue Aug 19 2025,@,Washington Mystics,W,,80,69,7,27,W 1,
35,Thu Aug 21 2025,,Washington Mystics,W,,67,56,8,27,W 2,
36,Sat Aug 23 2025,@,Chicago Sky,W,,94,84,9,27,W 3,
37,Mon Aug 25 2025,@,New York Liberty,L,,79,81,9,28,L 1,
38,Wed Aug 27 2025,@,Dallas Wings,W,,101,95,10,28,W 1,
39,Sat Aug 30 2025,,Minnesota Lynx,L,,70,94,10,29,L 1,
40,Mon Sep 1 2025,,Atlanta Dream,L,,76,93,10,30,L 2,
41,Wed Sep 3 2025,@,Chicago Sky,L,,64,88,10,31,L 3,
42,Sat Sep 6 2025,,Phoenix Mercury,W,,87,84,11,31,W 1,
43,Mon Sep 8 2025,@,Atlanta Dream,L,,62,87,11,32,L 1,
44,Wed Sep 10 2025,,Atlanta Dream,L,,72,88,11,33,L 2", col_types = cols(Date = col_date("%a %b %d %Y")))


basketball <- janitor::clean_names(basketball) 



  
```

We imported the data as a csv file, and then used a command to make the date variable be recognized as date data. Following this we cleaned the names of the variables, so they would be easier to use going forward. 

A brief look at the data:

```{r}
kable(basketball %>%
        head(n = 5))
```


## Exploratory Data Analysis

### Variable Exploration

#### Date Variable

```{r}
basketball %>%
mutate(wday = wday(date, label = TRUE)) %>%
ggplot(aes(x = wday)) +
geom_bar(color = "black",
         fill = "darkorange") +
  labs (x = "Day of the Week")
```

```{r}
basketball %>%
mutate(month = month(date, label = TRUE)) %>%
ggplot(aes(x = month)) +
geom_bar(color = "black",
         fill = "darkorange") 
```


#### Team Score

```{r}
basketball %>%
  summarize(
    Count = n(),
    Mean = mean(tm, na.rm = TRUE),
    Median = median(tm, na.rm = TRUE),
    sd = sd(tm, na.rm = TRUE),
    Min = min(tm, na.rm = TRUE),
    Q1 = quantile(tm, 0.25, na.rm = TRUE),
    Q3 = quantile(tm, 0.75, na.rm = TRUE),
    Max = max(tm, na.rm = TRUE)
  ) %>% 
  kable(caption = "Summary Statistics for Team Points")
```

```{r}
basketball %>% 
  ggplot(aes(x = tm)) +
    geom_histogram(bins = 15,
                   fill = "darkorange",
                   color = "black") +
    labs(title = "Distribution of Team Points (Histogram)",
         x = "Team Score",
         y = "Count")
```
```{r}
ggplot(basketball, aes(x=tm))+
  geom_boxplot() +
  labs (title = "Team Score Boxplot", 
    x = "Team Score")

```

#### Bivariate

```{r}
ggplot(basketball, aes(x = date, y = tm )) +
  geom_point(alpha = 0.2)+ 
  geom_smooth(color = "darkorange") +
  labs(x = "Date", y = "Team Score")
```



